# Most Dangerous Weather Phenomena in the USA

In this analysis, we will investigate which weather phenomena are

 * the most harmful to public health and which have
 * the greatest impact on the economy
 
across the whole of the United States of America.
We base our analysis on data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.

We investigate the total direct economic damage (in USD), the number of fatalities, and the number of injuries.
We order event types by the respective effect sizes, first summarized over the whole time of measurement but
also over a shorter timeframe.



## Data Processing

First we load the data set.
Documentation can be found 
  [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf);
there is also a much shorter
  [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf).

```{r loading data, cache=T}
if ( !file.exists("data.csv.bz2") ) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data.csv.bz2")
}
rawdata <- read.csv("data.csv.bz2")
```

For this analysis, we will focus on

 * (begin) date,
 * the number of fatalities,
 * the number of injuries, and
 * damages to property and crops.
 
In particular, we will ignore other related costs and delayed damage, as these data are only available in narrative form.

```{r selecting data}
library("dplyr")
data <- select(rawdata, REFNUM, EVTYPE, BGN_DATE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) 
```

Now we have some cleaning up to do.

 1. Damages are given in different units which are provided in the `_EXP` variables.
    We normalize the `_DMG` variables to dollars.
    
    We find the following values in `_EXP`:
    
    ```{r}
    library("knitr")
    summary(data$PROPDMGEXP)
    summary(data$CROPDMGEXP)
    ```
    
    We will interpret `k` and `K` as thousands, `m` and `M` for millions, and `B` as billions.
    If no value is given, we will interpret this as the amount given in plain dollars.
    Since this accounts for the overwhelming majority of observations we will ignore the other
    variants and put `NA`.
    
    Since we are only interested in overall damages, we just add the two different damage types.
    
    ```{r calc dmg}
    exponent <- function(e) {
      if ( e == "" ) {
        return(1)
      }
      else if ( e == "k" | e == "K" ) {
        return(1000)
      }
      else if ( e == "m" | e == "M" ) {
        return(1000000)
      }
      else if ( e == "b" | e == "B" ) {
        return(1000000000)
      }
      else {
        return(NA)
      }
    }
    
    data <- data %>% 
            rowwise() %>% # Need to go into rowwise mode because exponent takes only scalars
            mutate(DMG = PROPDMG * exponent(PROPDMGEXP) + CROPDMG * exponent(CROPDMGEXP)) %>%
            select(REFNUM, EVTYPE, BGN_DATE, FATALITIES, INJURIES, DMG) %>%
            ungroup() # remove effect of rowwise()
    ```
    
 2. We extract years from the (begin) dates; we do not need higher resolution.
 
    ```{r extract year}
    data <- data %>%
            mutate(YEAR = strftime(strptime(BGN_DATE, format = "%m/%d/%Y %H:%M:%S"), format = "%Y")) %>%
            select(-BGN_DATE)
    ```
    
Now we have the data in a form we can work with.

## Exploratory Analysis

In order to get a feeling of how the different effects distribute over the event types, we plot some quick histograms.

```{r summarize types}
evt_data <- data %>%
            group_by(EVTYPE) %>%
            summarize_each(funs(sum), FATALITIES, INJURIES, DMG)

summary(evt_data$FATALITIES)
summary(evt_data$INJURIES)
summary(evt_data$DMG)

library("ggplot2")
require(gridExtra)

hist1 <- ggplot(evt_data, aes(FATALITIES)) + geom_histogram()
hist2 <- ggplot(evt_data, aes(INJURIES)) + geom_histogram()
hist3 <- ggplot(evt_data, aes(DMG)) + geom_histogram()
grid.arrange(hist1, hist2, hist3, ncol=3)
```

It does indeed seem to be the case that very few event types dominate each effect.
We will therefore restrict ourselves to reporting the top 10 for each effect.

## Results

```{r sorting}
evt_by_fatal <- evt_data %>% 
                select(EVTYPE, FATALITIES) %>%
                mutate(FATALITIES_RATIO = FATALITIES / sum(FATALITIES)) %>%
                arrange(desc(FATALITIES))
evt_by_injur <- evt_data %>% 
                select(EVTYPE, INJURIES) %>%
                mutate(INJURIES_RATIO = INJURIES / sum(INJURIES)) %>%
                arrange(desc(INJURIES))
evt_by_damag <- evt_data %>% 
                select(EVTYPE, DMG) %>%
                filter(!is.na(DMG)) %>%
                mutate(DMG_RATIO = DMG / sum(DMG)) %>%
                arrange(desc(DMG))
```

Let us inspect the Top 10 for each effect. These lists directly tell us which phenomena have had the greatest total effect.
We note that, indeed, relatively few event types account for most of the damages done.

### Fatalities

```{r top fatalities}
kable(head(evt_by_fatal, 10), digits=2)
```

### Injuries

```{r}
kable(head(evt_by_injur, 10), digits=2)
```

### Economic Damage

```{r}
kable(head(evt_by_damag, 10), digits=2)
```

### Development over time

However, if we inspect the effects over time, we note that things may have been changing.

```{r final plots}
top_fatal_data <- data %>%
                  filter(EVTYPE %in% head(evt_by_fatal, 6)$'EVTYPE') %>%
                  group_by(EVTYPE, YEAR) %>%
                  summarize_each(funs(sum), FATALITIES)
top_injur_data <- data %>%
                 filter(EVTYPE %in% head(evt_by_injur, 6)$'EVTYPE') %>%
                 group_by(EVTYPE, YEAR) %>%
                 summarize_each(funs(sum), INJURIES)
top_damag_data <- data %>%
                 filter(EVTYPE %in% head(evt_by_damag, 6)$'EVTYPE') %>%
                group_by(EVTYPE, YEAR) %>%
                 summarize_each(funs(sum), DMG)

plot1 <- ggplot(data = top_fatal_data) +
  geom_point(aes(YEAR, FATALITIES, group = 1, color = EVTYPE, shape = EVTYPE))
plot2 <- ggplot(data = top_injur_data) +
  geom_point(aes(YEAR, INJURIES, group = 1, color = EVTYPE, shape = EVTYPE))
plot3 <- ggplot(data = top_damag_data) +
  geom_point(aes(YEAR, log(DMG), group = 1, color = EVTYPE, shape = EVTYPE))

# Equalize widths
g1 <- ggplotGrob(plot1)
g2 <- ggplotGrob(plot2)
g3 <- ggplotGrob(plot3)
g2$widths <- g1$widths
g3$widths <- g1$widths

grid.arrange(g1, g2, g3, nrow=3)
```

We make some observations.

 1. Economic damages seem to be spread out quite evenly. The top list above seems to be meaningful.
 
 2. For fatalities and injuries, only Tornado and TSTM Wind events were recorded before the 1990s (among the top 6). 
    Therefore, total sums over the whole date range are likely to misrepresent the danger of the respective
    types of events.
    
    Indeed, if we restrict ourselves to data since 1990, we obtain this:
    
    ```{r recent years}
    recent_data <- data %>%
                   filter(as.integer(YEAR) >= 1990) %>%
                   group_by(EVTYPE) %>%
                   summarize_each(funs(sum), FATALITIES, INJURIES, DMG)
    
    recent_by_fatal <- recent_data %>% 
                select(EVTYPE, FATALITIES) %>%
                mutate(FATALITIES_RATIO = FATALITIES / sum(FATALITIES)) %>%
                arrange(desc(FATALITIES))
    recent_by_injur <- recent_data %>% 
                select(EVTYPE, INJURIES) %>%
                mutate(INJURIES_RATIO = INJURIES / sum(INJURIES)) %>%
                arrange(desc(INJURIES))
    
    kable(head(recent_by_fatal, 10), digits=2)
    kable(head(recent_by_injur, 10), digits=2)
    ```
    
    While tornadoes still account for most injuries, fatalities are caused by a more diverse set of phenomena.
    
It is noteworthy that those events that cause the most damage to life and health do not cause the most economic damage, 
and vice versa.