---
title: "Data Science Specialization Course Project 6 -- Part 1"
author: "Raphael Reitzig"
date: "September 24, 2016"
output: pdf_document
---

## Part 1: A Simulation Exercise

We will need a bunch of exponentially random numbers.

```{r data}
lambda <- 0.2
expdata <- matrix(rexp(40000, lambda), ncol=40)
```

Now we will compute the set of sample means and variances we are interested in:

```{r stats}
means <- apply(expdata, 1, mean)
variances <- apply(expdata, 1, var)
```

### Distribution of the Sample Mean

Let us compare the sample means with the theoretical mean `r 1/lambda`.

```{r meanplot, out.width="0.6\\textwidth"}
library(ggplot2)
ggplot(data.frame(means=means), aes(means)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 1/lambda, color="dark red")
```

We see that we get a quite normal-looking distribution around the true mean.
With a standard deviation of `r sd(means)` it is not the most concentrated, though.


### Distribution of the Sample Variance

Let us compare the sample variances with the theoretical variance `r 1/(lambda*lambda)`.

```{r varplot, out.width="0.6\\textwidth"}
ggplot(data.frame(variances=variances), aes(variances)) +
  geom_histogram(bins = 30) +
  geom_vline(xintercept = 1/(lambda*lambda), color="dark red")
```

We see that we get a non-normal-looking distribution around the true variance.


### Comparison With Normal Distribution

We inspect how close the distribution of the mean is to normal. We superimpose the normal density using sample mean and standard deviation as parameters -- our best estimate for the distribution of the mean, if it *is* indeed normal -- onto the (smoothed and normalized) histogram.

```{r cmpmeanplot, out.width="0.6\\textwidth"}
estimated_normal <- function(x) { dnorm(x, mean = mean(means), sd = sd(means)) }

ggplot(data.frame(means=means), aes(means)) +
  geom_density(color="blue") + # This is a smoothed, normalized version of the histogram
  stat_function(fun = estimated_normal, color="dark red") # estimated distribution
```

The means seem indeed to be roughly normally distributed.