---
title: "Data Science Specialization Course Project 6 -- Part 2"
author: "Raphael Reitzig"
date: "September 24, 2016"
output: pdf_document
---

## Part 2: Tooth Growth

First, we need the data, of course.

```{r loadtoothdata}
data("ToothGrowth")
str(ToothGrowth)
```

We have 60 observations of lengths of certain cells of guinea pigs that were given different dosages of vitamin C as either orange juice (`OJ`) or in pure form (`VC`).

```{r toothplot, out.width="0.5\\textwidth"}
library(ggplot2)
ggplot(ToothGrowth, aes(x = factor(dose), y = len, fill = supp)) +
  geom_boxplot()
```

We want to show that higher dosage of vitamin C leads to longer cells for both forms of supplement.
To this end, we assume that the individual observations are i.i.d. normal (within each supplement-dosage group)
and that there are no confounding factors;
we can therefore apply the (grouped, unpaired, one-sided) t-tests. 

Our null-hypothesis is that the respective means are equal;
the alternative is that higher dosage results in greater cell length.
We seek significance of alpha = 0.05.

We compute p-values for the four relevant combinations.

```{r tests}
oj1 <- t.test(ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1, "len"],
              ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 0.5, "len"],
              alternative = "greater")
oj2 <- t.test(ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 2, "len"],
              ToothGrowth[ToothGrowth$supp == "OJ" & ToothGrowth$dose == 1, "len"],
              alternative = "greater")
vc1 <- t.test(ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1, "len"],
              ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 0.5, "len"],
              alternative = "greater")
vc2 <- t.test(ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 2, "len"],
              ToothGrowth[ToothGrowth$supp == "VC" & ToothGrowth$dose == 1, "len"],
              alternative = "greater")

results <- matrix(c(oj1$p.value, vc1$p.value, oj2$p.value, vc2$p.value), ncol=2)
rownames(results) <- c("Orange Juice", "Ascorbic Acid")
colnames(results) <- c("1 vs 0.5", "2 vs 1")
library(knitr)
kable(results)
```

We see that all p-values are less than alpha, that is we can reject the null hypothesis in all cases.

We conclude that, as far as this data goes, more vitamin C seems to cause larger cells in the present range 
of dosages and for the two forms of supplement we have seen here.